---
title: "Assignment_2.2"
author: "Austin Poole & Zach Hill"
date: "September 4, 2019"
output: word_document
---

# Course Project: Milestone 1

As far as science can tell, water is a requirement for life. The human animal itself is comprised mostly of water and with it comes the necessity of an environment capable of providing that water. With an ever-increasing population of humans the water needs of humanity are becoming greater year by year and that burden is felt by the planet more and more. At some point the toll being levied against our groundwater stores will exceed the replacement rate from precipitation in many areas leaving a lack of water not only for humanity but also for the crops required to support the human population and the surrounding flora and fauna.

The United States Geological Survey (USGS) tracks ground water supplies both historically and in real time. This data can be analyzed to help predict if our current groundwater supplies are adequate or if we are in danger of depleting groundwater stores due to overuse. This information can be used to aid in health and safety matters regarding safe and available water as well as in maintaining population control when water thresholds are exceeded. 

USGS records groundwater data through the Long-Term Groundwater Data Network which is comprised of " actively-measured periodic, continuous, and/or real-time wells with at least 20 years of measurements" (United States Geological Survey, 2019). The data available provide the option for 20, 30 and 50 year record samples from among over 9510 daily, monthly and annually recorded wells. Using historical data to predict trends in future water usage the federal, state and local governments can work together in forming a plan for the time when water will become limited. Due to the number of records available, the longer term records are generally only available from annual samples. The more recent data include a larger volume of samples in monthly and daily increments which will aid in fine-tuning usage allowances and requirements. Due to the volume of data collected in more recent years, analysis for generalized prediction will be accomplished using the 50 year dataset in annual increments.

The aim of this project is to help identify which areas in the Mid-Atlantic Hydrologic Region are at risk and when areas not already at risk might become so. Through anaylsis there should be insight gained regarding both our present situation, if there is one, and any concerns we might have for the future.

## Preliminary Requirement

Data must first be obtained for the USGS through https://groundwaterwatch.usgs.gov. The number of variables available for download is staggering so close attention must be paid when selecting relevant data. For this project, simple water elevations were selected to watch trends in water availability. These elevations are based on the North American Vertical Datum of 1988 (NAVD 88) survey standards. This multi-national survey standard set the height of the tidal benchmark for the continent and helped establish the mean sea level from which most modern measurements are made (National Geodetic Survey, 2018). 

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(reshape2)
library(tidyr)

sL <- list(380457075122301,
           380457075122302,
           380512075125401,
           380512075125402,
           380730075110401,
           380730075110402,
           380731075105601,
           380731075105602,
           380731075111101,
           380731075111102,
           380734075111301,
           380734075111302,
           380837075112201,
           380837075112202,
           381155075091801,
           381155075091802,
           381157075092901,
           381157075092902,
           381427075081102,
           381428075081401,
           381428075081402,
           381428075081403,
           381450075075401,
           381450075075402,
           381452075075901,
           381452075075902,
           381452075080101,
           381452075080102,
           381804075063401,
           381804075063402,
           381805075063501,
           381805075063502,
           381917076035501,
           381922076043001,
           381922076043002,
           381923076043501,
           381924076043801,
           381924076043901,
           381925076043001,
           381926076043401,
           381929076043001,
           381929076043002,
           382718076062002,
           383100076061101,
           383100076061103,
           383112076053503,
           383122076055701,
           383122076055702,
           390657076462601,
           390657076462602,
           390756076464201,
           390756076464202,
           390826076454801,
           390826076454802,
           390830076473902,
           392521075583301,
           392521075583302,
           392521075583303,
           392537075593001,
           392537075593002,
           392548075593802,
           392556075590001,
           392556075590002,
           431357073425301)

```
### Importing Data to R

After obtaining the site data, the dataRetrieval package in R was used to request the groundwater level data from each site and added to a dataframe for anaylsis. 

```{r load, messages = FALSE}
data <- dataRetrieval::readNWISgwl(sL, 
                                   startDate = "1900-01-01", 
                                   endDate = "2019-09-01")

df <- data.frame(data$site_no, data$lev_dt, data$lev_va)
```
### Data Structure

Multiple variables were included in the data retrieval process but of primary concern are the lev_va variable, which is the level above mean sea level (MSL) and the date/time stamps in this analysis. Additionally, site identifiers were used in case further reference is needed, especially for geographical positioning lookup.  The site numbers were imported as factors, the lev_dt variable is imported as Date format, and lev_va is imported as numeric. 
```{r sructure}
str(df)
```

### Possible Issues

Another hurdle might be consistency of readings as many wells levels were not recorded the full duration of the timespan to be analyzed. This might require a manipulation of the sites using generalized areas where wells are found if geographical coordinates can be obtained. 

## Technical Approach

### Analysis

```{r summary}
summary(df)
```
While the dataset requested should have contained results at least 50 years prior, the earliest observation with a timestamp was 1975. More recent dates seem to occur more frequently, as expected. One site in particular was very consistent, site number 381427075081102, with over 450 level readings. This comprises around 20% of the total data which will make for reliable analysis in that area but other regions aren't as consistent. 

### Requirement Development

Due to the nature of the hypothesis the first step in analysis is finding how the readings at each site relate to the other readings from that same site. This must be done prior to applying analysis to the region. Our first dataset will require reshaping to allow for better analysis and for some initial time-series visualization.


```{r time_series_df, echo = FALSE, messages = FALSE, warnings = FALSE}
ts_df <- dcast(df, data.lev_dt ~ data.site_no, value.var = 'data.lev_va', fun.aggregate = mean)

# Maybe we narrow down points with more than 10 years? 65 lines might be too much. Or maybe we just ditch wells with less than 5 years?

```
```{r time_series_window, echo = FALSE}
head(ts_df)[0:4]
```

This leaves us with a dataframe containing mostly NaN's due to most wells not reporting most dates. Even if a well reported within the same month, a different date will be reported so we will have to clean this up.

### Model Deployment

We will be starting with a time-series analysis of wells either in generalized areas or using the wells with the most consistently available readings in as many regions as possible. We should be able to use linear regression to predict outcomes based on the latter. As for the former, knn will need to be implemented to replace non-existent data in wells that did not report but which had nearby wells that did. The geographic well data does not exist in this dataset so we will need to find another source of data for this, which should also be available from USGS.

### Testing and Evaluation

Historical data may be used to verify our models with with known outcomes, then implemented on the non-training set to determine validity. It is expected that the data should match or provide insight into why it did not. 

## Expected Results

We expect to find that groundwater is depleting faster than it is replenshing. This is based on first hand observations of local water-level conditions around Charlotte, North Carolina. The area is a fast-growing population in a water-heavy area and should be a good indicator of how less-than-wet areas might be faring under the same population conditions.

## Management Plan

### Project Plan

After evaluation, regradless of outcome, results can be published to local lawmakers and media sources to inform residents of the current state of the local environment. This could have drastic impact were it to be found that the rise in population was putting the local farms at risk. 

### Project Risk

There is a chance it could be found to be correct, that the water levels are not replenishing as fast as they should be or that they have been in decline for some time and the public has not been made aware. There are many business proposals in the area currently for multi-billion dollar development deals that could be stymied were the people to find out they might not have water once these homes are filled. 